On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-04-27 Manuscript/manuscript.tex
removed or commented out fixed comments

\title{On Devon Allen's Disqualification at the 2022 Track and Field World
Championship}

@@

\jy{Avoid references in abstract.}

\jy{This is only the background. An abstract also needs to summarize the 
contributions.}
%\bigskip
\eds{removed Devon Allen from keywords because his name is already in title
and replaced with others}

\noindent\sc{Keywords}:
False Start, Hurdles, Reaction Time, Seiko, 2022 World Track and Field Championships

@@

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championship.
\eds{you mention in the final section that the next world championship is in 
2023, so he won't need to wait two years if that is true}
\of{Resolved}
%\eds{final two sentence seems out of place here; consider moving elsewhere?}
%\of{Moved to the end of paragraph 4 of introduction}

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championship to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged \eds{can we use different word 
instead of wrong?  at fault?  However, should that be 0.05\% chance at fault?}
\of{Changed to "wronged"}
came from Johnson claiming there is a $(1/13)^3$ that the fastest reaction
times came during the same championship for the men's 100 meter,
men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 
2022 \eds{?} \of{Cleaned up but may need additional work}
\citep{johnson2022was}.
%\eds{compared the reaction times at the 2022 World Track and Field  Championship 
%to what? Also restate here that no formal statistical analyses have been done.}
%\of{Cleaned this up a little but still a little messy.}

@@

\jy{Could be condensed}
\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 Track and Field World Championship 
and at his physical peak.

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). \eds{are you using covariates?  Data section
implies that you are, but Results section does not}
\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.
\eds{capitalize first letter when referring to a numbered section, table, figure, 
etc. ``Section 4 provides...'' vs. ``The next section provides...''}
\of{I think capitalization should be all set: all are in form of ``Section i''}

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championship or a 2022 national competition held within two months of the  
2022 World Championships). The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank-based comparisons.

\subsection{World Championship}\label{sec:dataworld}

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
  \label{fig:Boxplots}
\end{figure}

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; and can be found in
the Appendix~\ref{sec:Appendix}.   \eds{see earlier comment - 
are you using any of these as covariates?}
\of{removed covariate confusion, talk about womens a couple sentences down} 
We chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}
The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
semifinal heat (of which there are either two
or three each year) denoted by ``S", or final heat denoted by ``F".
The preliminary heats were not studied because athletes react faster as the 
competition progresses \citep[e.g.,][]{zhang2021correlation}. The variable 
``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  The
variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly). There is no statistical meaning to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
\eds{if you are really not using ``H'' data at all, it does not appear necessary
to spend time defining earlier.  You could just say directly when you define 
stage that you only consider semifinals and finals and cite the Zhang paper for 
rationale}
\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few finals observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.
The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both 
the men's finals and semifinals data is lower (indicating faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found median reaction times of 0.156 seconds \eds{mean?  median?} 
\of{Resolved}
for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
This was desirable for World Athletics as false starts can make already long 
track meets more tedious, both for athletes and viewers watching the television 
broadcast.  There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds \eds{for men, not overall} \of{Better?} in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championship there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.
\eds{is that right?  it seems counterintuitive to me}
\of{Yes you are correct, I updated the language.}

\eds{I am not sure the previous sentence follows from the one preceding it. Why
 does World Athletics find faster reaction times desirable?  Do you mean that
they find fewer false starts desirable?}

\eds{Two comments about the previous sentence: 1. 
it does not logically follow the previous sentence as there is no mention of 
gender, and 2. if you are not using women's data, don't spend time talking about 
it at the beginning of the data section.  Similar to my comment about 
preliminary heats, you can move this sentence to the beginning of the Data 
section to provide a rationale for why you do not include data for women.} 
\of{Updates: I moved any talk about women's data to start of Data section, I think
it flows much better now}

\subsection{Beyond World Championship}\label{sec:databeyond}
The motivation for looking beyond World Championship data came from an 
exploratory \eds{exploratory instead of qualitative?} \of{I agree; fixed}
analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
\eds{be consistent with how you refer to the US championship:
``United States Track and Field Championships" and ``United States Track and Field 
Championships" are both being used; please pick one and use that throughout}
\of{Resolved; changed all USA to United States.}
%This section was motivated from a qualitative analysis that started by looking
%at how United States athletes reacted at the 2022 United States Track and Field
%Championships. 
From June 23-26 2022, the United States held its Track
and Field Championship at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the United States Track and Field 
Championship races. Devon Allen raced three times at the United States Track and
Field Championship and three times at the World Track and Field Championship. 
His reaction times were substantially faster in the World Track and
Field Championship compared to the United States Track and Field Championship as were his
teammates. \eds{statistically significantly faster? If you used a statistical 
test, it should not be described in this section.  If you didn't use a 
statistical test, you should replace the word `significantly' with something 
else (substantially?)} \of{Replaced signficantly with substantially}
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``United States" denotes
%the United States Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

@@

\of{Im not sure how to fix this but the spacing between the lines of the
table is different than the rest of the paper}

@@

The %World Championship data and the data eyond the World Championship 
data described in Sections~\ref{sec:dataworld} and~\ref{sec:databeyond} were
analysed with GLMM and rank-based methods, respectively.
\eds{need consistency with ``Championship" or ``Championships - I am not sure 
which one is correct}

\subsection{GLMM}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championship, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
how is this athlete clustering being accounted for here?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
Gamma family with an identity link works well for their reaction-time data.
It doesn't seem like a good idea in general because it is likely that some 
negative predicted responses will occur either during model fitting or in the 
final model, but the internet seems to favor this link so a reviewer may ask 
about it too.}
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

@@

As shown in table~\ref{tab:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study~\ref{subsec:Results_Rank}.

>>>>

\title{On Devon Allen's Disqualification at the 2022 World Track and Field 
Championships}

@@

%\jy{Avoid references in abstract.}

\jy{This is only the background. An abstract also needs to summarize the 
contributions.}
%\bigskip
%\eds{removed Devon Allen from keywords because his name is already in title
%and replaced with others}

\eds{also removed 2022 World Track and Field Championships from keywords and 
changed location of World in the title. World Track and Field Championships
has more Google search results than Track and Field World Championships, so
I opted to go with this. Also used plural Championships. Feel free to switch 
order or use singular if one or both of these is 
preferred for some reason.  }

\noindent\sc{Keywords}:
False Start, Hurdles, Reaction Time, Seiko

@@

%1)
Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
%After coming in third in the 110 meter hurdles at the United States Track and 
%Field Championships also in Eugene at the end of June, Allen was poised to 
%compete in the 2022 World Track and Field Championships also 
%located at Eugene in August.
%two months later. 
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 in his preliminary heat and 0.101 in his semifinal.  
%Allen, who attended the University of Oregon also in Eugene, 
%(the location of US World Championship 
%\eds{US World? remove World?} and the World Championship) 
%was
%considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
%in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
%Thus 
His disqualification in the final heat by reacting in 0.099 seconds instead of 
0.1 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification; it can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers 
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

%2)
Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championship and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championship to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same championship 
\citep{johnson2022was}.
%came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
%reaction times came during the same championship for the men's 100 meter,
%men's 110 meter hurdle, and women's 100 meter for 13 competitions held between 2003 and 2022 
%\of{Cleaned up but may need additional work}
\of{Cleaned this up a little but still a little messy.}
\eds{is this better? (and still correct?)}

@@

\jy{Could be condensed}
\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}. This 
contradicts the results of \citet{zhang2021correlation}, which indicated that 
year of competition has a significant effect on reaction times.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.001 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 World Track and Field Championships 
and at his physical peak.

@@

%4)
\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  First we model the data, which is
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with a random effect 
accounting for the venue (or year). 
%\eds{are you using covariates?  Data section
%implies that you are, but Results section does not}
%\of{Cleaned up confusion about covariates, but no none were used in the paper} 
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 data differed significantly from other 
competitions.
Our investigation into reaction times is important not only because if its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.
%\eds{I recommend moving the rest of this paragraph to Section 2.2 as I don't
%think it is critical for readers to know all the details here.}
%First we can compare the reaction times of athletes who
%competed at the World Track and Field Championship in both 2019 and 2022, and 
%second we can compare how athletes performed at the 2022 championship relative 
%to their national/qualifiers meet. In order to reduce any chances
%of athletes improving reaction time, we can also look at data from qualifying meets.
%Almost every athlete had to qualify by proving they are elite at the national level, 
%usually by coming in at least fourth. Thus, data was compiled on athletes who competed
%at the 2022 World Track and Field Championships on what these athletes also ran 
%at their country's national championship. If there was any argument that athletes
%have improved their reaction time from 2019 to 2022, it should now be considered
%nearly null, as the national competitions took place between June and July 2022 
%and the World Championships were in August. The equipment used to measure reaction
%time cannot be guaranteed to be the same from national to international
%competition, but the time span between meets has been reduced considerably.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

\subsection{World Championships}\label{sec:dataworld}

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \caption{The reaction times from 1999 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
  \label{fig:Boxplots}
\end{figure}

Data was taken from the World Athletics 
%\eds{is there a website you can cite here?}
and covers the men's 110 meter hurdles from 1999 to 2022; the source can be 
found in the Appendix.%~\ref{sec:Appendix}.   %\eds{see earlier comment - 
%are you using any of these as covariates?}
%\of{removed covariate confusion, talk about womens a couple sentences down} 
Although women's data is available, we chose not to include women's data in this
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}.
%The variable ``Stage" refers to whether the observation occurred during a 
%heat (preliminary round)  
%semifinal heat (of which there are either two
%or three each year) denoted by ``S", or final heat denoted by ``F".
We focus on the reaction times during the semifinal and final heats only, as 
reaction times from preliminary heats are often not as fast as those in later 
heats %were not studied because athletes react faster as the 
%competition progresses 
\citep[e.g.,][]{zhang2021correlation}. 
%The variable 
%``Gender" has levels ``M" and ``F" for ``male" and ``female" respectively.  
%The variable ``Batch" takes a numerical value and every heat of every round from 
%1999 to 2022 was assigned it's
%own number.  This was added because of suspected issues that reaction time may be
%correlated with other runners reacting quickly (one runner reacting quickly may
%result in other runners reacting quickly). There is no statistical meaning to
%how the batches were labeled, they were ordered chronologically by meet but
%reverse chronologically by year (Batch 1 is a preliminary heat of 2022 and Batch
%111 corresponds to the 1999 finals). %Included are
%box plots of the reaction times of the men's finals and the men's semifinals and
%finals pooled together.  
For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together and refer to this data as the ``pooled data"
%We will refer to the latter as the ``pooled data"
throughout the paper.  %The heats or preliminary rounds were not studied because
 % Athletes react 
%faster in the finals than the semifinals which are faster than the heats 
%\citep{zhang2021correlation}.  
%\eds{if you are really not using ``H'' data at all, it does not appear necessary
%to spend time defining earlier.  You could just say directly when you define 
%stage that you only consider semifinals and finals and cite the Zhang paper for 
%rationale}
%\of{Ok I updated this.  Should I delete the batch variable too?  Although I don't
%use it, I do reference it in the methods section}
The pooled data increases the sample size which is advantageous because in some
years there are very few final heat observations.  For example, in 2022 there
were only five data points due to two disqualifications and one athlete who 
did not compete.

The data under consideration is best visualized using side-by-side box plots, as
shown in Figure~\ref{fig:Boxplots}.  It is clear that the reaction times in the 
2022 box plot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\citet{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
This was desirable for World Athletics as false starts can make already long 
track meets more tedious, both for athletes and viewers watching the television 
broadcast.  There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championship there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.
  %\eds{I am not sure the previous sentence follows from the one preceding it. Why
% does World Athletics find faster reaction times desirable?  Do you mean that
%they find fewer false starts desirable?}

%\eds{Two comments about the previous sentence: 1. 
%it does not logically follow the previous sentence as there is no mention of 
%gender, and 2. if you are not using women's data, don't spend time talking about 
%it at the beginning of the data section.  Similar to my comment about 
%preliminary heats, you can move this sentence to the beginning of the Data 
%section to provide a rationale for why you do not include data for women.} 
%\of{Updates: I moved any talk about women's data to start of Data section, I think
%it flows much better now}

\subsection{Beyond World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.
%\eds{be consistent with how you refer to the US championship:
%``United States Track and Field Championships" and ``United States Track and Field 
%Championships" are both being used; please pick one and use that throughout}
%\of{Resolved; changed all USA to United States.}
From June 23-26 2022, the United States held its Track
and Field Championships at Hayward 
Field in Eugene, Oregon to decide who to send to the World Championships 
being held at the same venue in August. %the exact same venue as used 
%for the World Track and Field Championship a month later.  
%Thus, a baseline is able to be established for the four United States athletes who 
Thus, we can establish a baseline for the four United States athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all the World 
Track and Field Championship races compared to the United States Track and Field 
Championship races. Devon Allen raced three times at the United States Track and
Field Championship and three times at the World Track and Field Championships. 
His reaction times were substantially faster in the World Track and
Field Championships compared to the United States Track and Field Championships 
as were his
teammates. 
This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  Table~\ref{fig:USAvsWorld} highlights the difference in reaction 
times for Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. 
%``United States" denotes
%the United States Track and Field Championships, ``World" denotes the World Track and Field
%Championships. All numbers listed in the table are the reaction
%time for the respective athletes in seconds.

@@

\of{Im not sure how to fix this but the spacing between the lines of the
table is different than the rest of the paper}
\eds{LaTeX keeps captions singlespaced automatically with the doublespace 
command; don't worry about it!}

@@

The %World Championship data and the data eyond the World Championship 
data described in Sections~\ref{sec:dataworld} and~\ref{sec:databeyond} were
analysed with GLMM and rank-based methods, respectively.
\eds{need consistency with ``Championship" or ``Championships - I am not sure 
which one is correct.  I am using Championships (plural) but
change everything to singular if that is preferred}

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
\eds{we know that many of the athletes compete in multiple world championships,
how is this athlete clustering being accounted for here?}
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
\eds{Did you try an identity link also? Lo and Andrews (2015) suggest that the 
Gamma family with an identity link works well for their reaction-time data.
It doesn't seem like a good idea in general because it is likely that some 
negative predicted responses will occur either during model fitting or in the 
final model, but the internet seems to favor this link so a reviewer may ask 
about it too.}
We include a random effect for venue because the technology
that World Athletics has used has differed since 1999, and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art. Additionally, any other weather or climate related factors 
such as humidity, precipitation, elevation, could be captured by this effect.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

@@

As shown in Table~\ref{tab:Gamma_parameters}, the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).
